 

### 2021-04-15 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

In particular, hydraulic constraints increase with height on a tree [REFS; *McDowell-isotopes book chapter*; @cobleLightAcclimationOptimizes2015; *Couvreur et al., 2018*] such that gas exchange is strongly constrained by at the tops of(~100 m. tall *Sequoia sempervirens*; *Ambrose et al. 2009, 2010*).
Under constant light conditions (photon flux density), taller trees (>35m) have lower canopy stomatal conductance compared to shorter trees  [@schaferEffectTreeHeight2000; REF]. Adaptation to increasing height and hydraulic contraints show that leaf size [@chinLeafAcclimationLight2017] and lobation [@kusiPlasticLeafMorphology2020] tend to decrease, whereas leaf mass per area (at greater than 20m) has shown to increase [@cobleLightAcclimationOptimizes2015; @kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a].

@@

Leaf traits in woody species take approximately 30-60 days to stabilize to integrated light after formation [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. As a result, degrees of high to low irradiance gives rise to a dichotomy of sun and shade leaf traits. Typical sun leaves are often smaller in size and width, thicker, and hairier, with thicker cuticle and greater lobation than shade leaves [Rozendaal et al. 2006, Mathur et al. 2018, *REFS*]. Within-species, several studies point towards these common variations in leaf anatomical, structural and biochemical trait plasticity at a vertical intra-canopy gradient based on adaptation to primarily light conditions, and as well biophysical gradient [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.

However, even with distinct features of sun and shade traits, trait adaptation to light maximizes light use efficiency for photosynthesis at both high and low light, while also adapting to other stressors as mentioned above. In light limited conditions, shade leaves increase light capture efficiency by increasing their specific leaf area [SLA] and by investing in chlorophyll on a mass basis, which "leads to a similar chlorophyll content on an area basis compared with sun leaves. This enables shade leaves to realize similar light capture to sun leaves at a lower biomass investment" [Chazdon et al. 1996, Poorter et al. 2000, @rozendaalPlasticityLeafTraits2006]

An increasing number of studies point towards the importance of incorporating sun and shade in modelling leaf economic spectrum (LES) traits within-canopy gradient (???), because LES theory was developed using mainly sun exposed (upper canopy) leaves. Given the vastly different light environments at the bottom of the canopy and that most leaves in the canopy are likely to be shade leaves, there is a strong research need to better characterize trait relationships and leaf responses through the whole vertical canopy [Keenan and Niinemets 2016]. However, there is also concern for ambiguity in a simplified categorization of 'sun' and 'shade' leaves, due to variation based on within-canopy light, across functional groups, in canopy gaps, and with height and ontogeny (Niinemets et al., 2015; Ishii et al., 2018) as well as geographically (Ishii et al., 2018).

@@

>>>>

In particular, hydraulic constraints increase with height on a tree [REFS; *McDowell-isotopes book chapter*; @cobleLightAcclimationOptimizes2015; *Couvreur et al., 2018*] such that gas exchange is strongly constrained by at the tops of(~100 m. tall *Sequoia sempervirens*; *Ambrose et al. 2009, 2010*).
Under constant light conditions (photon flux density), increase in height (>35m) showed decrease in canopy stomatal conductance, along with increasing $VPD$  [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010]. Adaptations to increasing height and hydraulic contraints show that leaf size [@chinLeafAcclimationLight2017] and lobation [@kusiPlasticLeafMorphology2020] tend to decrease, whereas leaf mass per area (at greater than 20m) tends to increase [@cobleLightAcclimationOptimizes2015; @kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a; @chinLeafAcclimationLight2017].

@@

Leaf traits in woody species take approximately 30-60 days to stabilize to integrated light after formation [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. As a result, degrees of high to low irradiance gives rise to a dichotomy of sun and shade leaf traits. Trait adaptation maximizes light use efficiency for photosynthesis at differing light levels and conditions mentioned above. Therefore, typical sun leaves are often smaller in size and width, thicker, and hairier, with thicker cuticle and greater lobation than shade leaves, where as shade leaves increase light capture efficiency by increasing their specific leaf area [SLA] and by investing in chlorophyll on a mass basis, which "leads to a similar chlorophyll content on an area basis compared with sun leaves. This enables shade leaves to realize similar light capture to sun leaves at a lower biomass investment" [Chazdon et al. 1996, Poorter et al. 2000; @rozendaalPlasticityLeafTraits2006; Mathur et al. 2018]. Within-species, several studies point towards these common variations in leaf anatomical, structural and biochemical trait plasticity at a vertical intra-canopy gradient across individual trees [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.

An increasing number of studies point towards the importance of incorporating sun and shade in modelling leaf economic spectrum (LES) traits within-canopy gradient [Chen et al. 2019], because LES theory was developed using mainly sun exposed (upper canopy) leaves. Given the vastly different light environments at the bottom of the canopy and that most leaves in the canopy are likely to be shade leaves, there is a strong research need to better characterize trait relationships and leaf responses through the whole vertical canopy [Keenan and Niinemets 2016]. However, there is also concern for ambiguity in a simplified categorization of 'sun' and 'shade' leaves, due to variation based on within-canopy light, across functional groups, in canopy gaps, and with height and ontogeny (Niinemets et al., 2015; Ishii et al., 2018)

@@

Across ontogeny, while canopy and understory trees of same species in tropical, temperate, and conifer forests show significant differences in leaf traits, the difference has been largely attriuted to changes in light conditions [Rosendaal et al. 2006], therefore maintaing similar responses to changes in light with increasing height as highlight within species. Under open-light conditions for canopy and understory,in large gaps and open-grown conditons, 
evidence shows that canopy trees showed significant increase in LMA, photosynthetic capcity per area, and stomatal conductance compared to saplings
[Thomas and Winner, 2002]. However, plasticity to light has been shown difference between species [Houter and Pons 2012]
